---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# lbkeyboard

<!-- badges: start -->
<!-- badges: end -->

lbkeyboard provides several functions to generate optimized keyboard layouts for Luxembourg; this means that the layout
needs to work well with French, German, Luxembourguish and English. These 4 languages are the most used languages in the
country, with French being (very likely) the most used language.

## Installation

You can install the development version of lbkeyboard from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("b-rodrigues/lbkeyboard")
```

## Example

For now, this package only allows you to create a plot of several layouts.

```{r example}
library(lbkeyboard)
## basic example code
```

Here is the BÉPO layout:

```{r}
ggkeyboard(keyboard = afnor_bepo, layout = "iso")
```

and here the AZERTY layout:

```{r}
ggkeyboard(keyboard = afnor_azerty, layout = "iso")
```

AZERTY and BÉPO are the layouts used in France; AZERTY is, by a very wide margin, the most used layout in France.
BÉPO is an optimized layout for the French layout.

In Luxembourg, the most used layout is the Swiss French (or German, depending on the user's preferences):

```{r}
ggkeyboard(keyboard = ch_qwertz, layout = "iso")
```

It is also possible to generate heatmaps. First, let's load some French text:

```{r}
data("french")
```

Then, let's compute the frequency of the letters:

```{r}
letter_freq_df <- letter_freq(french)
```

Let's add the colours:

```{r}
heatmap_azerty <- heatmapize(afnor_azerty, letter_freq_df)
```

Let's plot the heatmap:

```{r}
ggkeyboard(heatmap_azerty)
```

Let's compare to the BÉPO layout:


```{r}
heatmap_bepo <- heatmapize(afnor_bepo, letter_freq_df)
ggkeyboard(heatmap_bepo)
```

Let's compare heatmaps for Swiss German (QWERTZ) and BÉPO for the German language:

```{r}
data("german")
letter_freq_de <- letter_freq(german)

heatmap_qwertz <- heatmapize(ch_qwertz, letter_freq_de)
heatmap_bepo_de <- heatmapize(afnor_bepo, letter_freq_de)

ggkeyboard(heatmap_qwertz)
ggkeyboard(heatmap_bepo_de)
```

Even though BÉPO is originally a layout optimized for the French language, we can see from the heatmap that the 
most frequently used letters are in the home row, which is not the case for the Swiss German (QWERTZ layout).

Let's also generate some heatmaps for English:


```{r}
data("english")
letter_freq_en <- letter_freq(english)

heatmap_qwertz <- heatmapize(ch_qwertz, letter_freq_en)
heatmap_bepo_en <- heatmapize(afnor_bepo, letter_freq_en)

ggkeyboard(heatmap_qwertz)
ggkeyboard(heatmap_bepo_en)
```

Here again, we see that the BÉPO layout seems to fare better than the QWERTZ layout.




## Optimization Algorithm

The package includes a genetic algorithm for finding optimal keyboard layouts. The algorithm uses a **Carpalx-inspired effort model** that considers:

- **Base key effort**: Based on row position, finger strength, and distance from home position
- **Same-finger bigrams**: High penalty when consecutive keys use the same finger
- **Same-hand sequences**: Moderate penalty, with inward rolls penalized less than outward rolls
- **Row changes**: Penalty for reaching between rows on same hand

### Basic Usage

```{r, eval=FALSE}
# Load sample texts
data(french)
data(luxembourguish)

# Optimize layout for multiple languages
result <- optimize_layout(
  text_samples = c(french, luxembourguish),
  generations = 200,
  verbose = TRUE
)

# View the optimized layout
print(result$layout)

# Check improvement over starting layout
cat("Improvement:", round(result$improvement, 2), "%\n")
```

### Comparing Layouts

You can compare the typing effort across different keyboard layouts:

```{r, eval=FALSE}
data(afnor_bepo)
data(afnor_azerty)
data(ch_qwertz)

comparison <- compare_layouts(
  keyboards = list(
    BEPO = afnor_bepo,
    AZERTY = afnor_azerty,
    QWERTZ = ch_qwertz
  ),
  text_samples = french
)
print(comparison)
```

### Effort Breakdown

Get detailed breakdown of effort components:

```{r, eval=FALSE}
breakdown <- calculate_layout_effort(
  afnor_bepo,
  french,
  breakdown = TRUE
)
print(breakdown)
```

### Fixed Key Positions

You can keep certain keys in their original positions while optimizing the rest:

```{r, eval=FALSE}
# Keep home row keys fixed, optimize everything else
result <- optimize_layout(
  text_samples = french,
  fixed_keys = c("a", "s", "d", "f", "j", "k", "l"),
  generations = 200
)

# Check how many keys were optimized
cat("Fixed:", result$n_fixed, "Optimized:", result$n_optimized, "\n")
```

This is useful for:
- Keeping frequently-used keys in familiar positions
- Preserving muscle memory for certain keys
- Partial optimization of existing layouts

### Algorithm Parameters

The genetic algorithm supports tuning:

- `population_size`: Number of candidate layouts (default 100)
- `generations`: Evolution iterations (default 500)
- `mutation_rate`: Probability of mutation (default 0.1)
- `crossover_rate`: Probability of crossover (default 0.8)
- `effort_weights`: Weights for different effort components
- `fixed_keys`: Keys to keep in their original positions

The algorithm is implemented in C++ via Rcpp for efficiency

## Acknowledgment

This package uses some code from my fork of the [ggkeyboard](https://github.com/b-rodrigues/ggkeyboard) package 
(original package can be found [here](https://github.com/sharlagelfand/ggkeyboard).
