---
title: "Introducing QWERTZ-LUX: A Better Keyboard Layout for Luxembourg"
author: "Bruno Rodrigues"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introducing QWERTZ-LUX: A Better Keyboard Layout for Luxembourg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4,
  cache = FALSE
)
library(lbkeyboard)
library(dplyr)
```

## TL;DR

I've been thinking about keyboard layouts for a while. QWERTZ isn't great, and France
recently adopted BÉPO as an official standard alongside a revised AZERTY. What would
such an optimized layout look like for Luxembourg? Here's my proposal: **QWERTZ-LUX**.

## The Problem with QWERTZ

If you've ever wondered why you're typing "E" with your middle finger stretched up
to the top row while "J" (which you barely use) sits comfortably on the home row...
well, you're not alone. The QWERTZ layout, like its QWERTY cousin, was designed in
the 1870s for typewriters. The primary concern wasn't typing efficiency—it was
preventing mechanical jamming of type bars.

**Fast forward 150 years**: we're still using essentially the same layout, even though
the mechanical constraints disappeared decades ago.

The home row—where your fingers naturally rest—should contain the most frequently
used letters. But look at QWERTZ's home row:

```
A S D F G H J K L
```

For a multilingual environment like Luxembourg (where we type in French, German,
Luxembourgish, and English), this means:

- **E** (the most frequent letter in all four languages) is on the top row
- **T, N, R, I** (all extremely common) are scattered or require reaching
- **J, K** (rarely used) take prime real estate on the home row

This is objectively inefficient.

## France Shows the Way

In 2019, France officially adopted two new keyboard standards:

1. **A revised AZERTY** that fixes many issues while maintaining familiarity
2. **BÉPO** — a completely redesigned layout optimized for French

I've been using BÉPO for years, and I can honestly say it's a game-changer. Typing
feels more natural, there's less finger movement, and the learning curve, while steep
at first, pays off significantly.

But here's the thing: BÉPO was designed primarily for French. Luxembourg is unique—
we need something that works for French *and* German *and* Luxembourgish *and* English.

## The QWERTZ-LUX Approach

So I started wondering: what would an optimized keyboard layout for Luxembourg look
like?

I had two options:

1. **Go full BÉPO** and adapt it for Luxembourgish needs
2. **Keep QWERTZ as a base** and optimize only what's necessary

I chose a middle path. QWERTZ-LUX is essentially a fork of BÉPO's structure, but
with modifications that:

- Keep familiar elements for QWERTZ users (Ctrl+Z, Ctrl+X, Ctrl+C, Ctrl+V shortcuts)
- Optimize the letter arrangement for our multilingual corpus
- Provide direct access to accented characters (é, à, ü, ö, ç, ä) without dead keys

The home row had to change substantially—QWERTZ's home row is simply too inefficient.
But I tried to minimize disruption elsewhere.

## How Much Better Is It?

Let's look at the numbers. Using the **lbkeyboard** package's effort model (which
accounts for finger travel, same-finger penalties, and ergonomic factors), here's
how the layouts compare on a typical multilingual Luxembourg corpus:

```{r load-corpus, echo=FALSE}
# Load sample texts and create weighted corpus
data("french")
data("german")
data("english")
data("luxembourguish")

corpus <- list(
  French = rep(french, 3),         # 30%
  English = rep(english, 3),       # 30%
  German = rep(german, 2),         # 20%
  Luxembourgish = rep(luxembourguish, 2)  # 20%
)

all_texts <- unlist(corpus)
freq_all <- letter_freq(paste(all_texts, collapse = " "))
```

```{r effort-comparison, echo=FALSE}
# Load layouts
data("ch_qwertz")
data("afnor_bepo")

# Create QWERTZ-LUX layout using the package function
qwertz_lux <- create_qwertz_lux_keyboard()

# Effort weights
effort_weights <- list(
  base = 3.0,
  same_finger = 3.0,
  same_hand = 0.5,
  row_change = 0.5,
  trigram = 0.3
)

# Calculate effort for each layout
effort_qwertz <- calculate_layout_effort(ch_qwertz, all_texts, 
                                         keys_to_evaluate = letters,
                                         effort_weights = effort_weights)

effort_bepo <- calculate_layout_effort(afnor_bepo, all_texts,
                                       keys_to_evaluate = letters,
                                       effort_weights = effort_weights)

# QWERTZ-LUX: include both letters and the direct accents
qwertz_lux_keys <- c(letters, "é", "ä", "ë", "ç")
effort_qwertz_lux <- calculate_layout_effort(qwertz_lux, all_texts,
                                             keys_to_evaluate = qwertz_lux_keys,
                                             effort_weights = effort_weights)

# Create comparison table
comparison_df <- data.frame(
  Layout = c("QWERTZ (Swiss)", "BÉPO", "QWERTZ-LUX"),
  Effort = c(effort_qwertz, effort_bepo, effort_qwertz_lux),
  stringsAsFactors = FALSE
) %>%
  mutate(
    `Relative (%)` = round(Effort / min(Effort) * 100, 1),
    `vs QWERTZ` = paste0(round((1 - Effort / effort_qwertz) * 100, 1), "%")
  ) %>%
  arrange(Effort)

comparison_df$Effort <- round(comparison_df$Effort, 2)

knitr::kable(
  comparison_df,
  caption = "Typing effort comparison (lower is better)",
  col.names = c("Layout", "Effort Score", "Relative (%)", "Improvement vs QWERTZ"),
  align = c("l", "r", "r", "r")
)
```

## The QWERTZ-LUX Layout

Here's what the optimized layout looks like:

```{r show-layout, echo=FALSE, results='asis'}
# Display layout in text format grouped by row
cat("**Base Layer (normal keys):**\n\n")
cat("```\n")

# Get keys by row from qwertz_lux data frame
for (r in sort(unique(qwertz_lux$row))) {
  row_keys <- qwertz_lux[qwertz_lux$row == r, ]
  row_keys <- row_keys[order(row_keys$number), ]
  cat(paste(toupper(row_keys$key), collapse = " "), "\n")
}
cat("```\n\n")

cat("**AltGr Layer (additional accents via AltGr):**\n\n")
cat("```\n")
cat("è  à  ü  ö  (and more via dead keys: ê, â, ô, î, û, ï)\n")
cat("```\n")
```

### Key Features

| Feature | Benefit |
|---------|---------|
| **Optimized home row** | High-frequency letters (E, T, N, R, I, S) within easy reach |
| **Direct accents** | é, à, ü, ö, ç, ä accessible without dead key combinations |
| **ZXCV preserved** | Common shortcuts (Ctrl+Z/X/C/V) stay in familiar positions |
| **BÉPO-compatible** | Same modifier layer philosophy as BÉPO |

## Should You Switch?

That depends on your priorities:

- **If you value efficiency**: Yes. The numbers don't lie—QWERTZ-LUX requires
  meaningfully less effort for the same typing tasks.
  
- **If you're a touch typist**: You'll need to relearn the layout, but the
  investment pays off. I'd estimate 2-4 weeks of practice to regain full speed.
  
- **If you type casually**: The switch might not be worth the learning curve.
  But if you're reading this, you're probably not a casual typist!

## What's Next?

I'm still refining this layout based on real-world usage. The **lbkeyboard** package
lets you:

- Analyze your own text corpora
- Experiment with different layout modifications
- Quantify improvements with the effort model

Want to try it? Install the package and dive in:

```{r install, eval=FALSE}
# Install from GitHub
# remotes::install_github("b-rodrigues/lbkeyboard")

library(lbkeyboard)

# Explore the layout
qwertz_lux <- parse_layout_ini("layouts/qwertz-lux/layout.ini")
print(qwertz_lux)

# Visualize it
plot_layout_ini("layouts/qwertz-lux/layout.ini")
```

---

## Discussion

I'd love to hear your thoughts:

1. **What letters/symbols do you type most often?** (This helps tune the model)
2. **What shortcuts can't you live without?** (I tried to preserve common ones)
3. **Would you try an optimized layout?** (Or are you QWERTZ-for-life?)

The beauty of having this as an R package is that *you* can fork it and create
your own optimized layout for your specific needs. The code is open source, the
methodology is transparent, and the effort model is documented.

Happy typing! 🎹

---

## Session Info

```{r session-info}
sessionInfo()
```
