 ---
title: "QWERTZ vs Optimized QWERTZ: A Comparison"
author: "lbkeyboard"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{QWERTZ vs Optimized QWERTZ: A Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4,
  cache = FALSE
)
# For development: load_all() ensures latest code is used
if (file.exists("../DESCRIPTION")) {
  devtools::load_all("..")
} else {
  library(lbkeyboard)
}
library(dplyr)
library(ggplot2)
```

## Introduction

This vignette provides a focused comparison between the **standard QWERTZ** layout 
(commonly used in Switzerland, Germany, and Luxembourg) and an **optimized QWERTZ** 
variant with an improved home row, designed specifically for Luxembourg's multilingual 
environment.

We will:

1. **Compare typing efforts** between the two layouts
2. **Visualize with heatmaps** where keystrokes concentrate
3. **Analyze accented character frequencies** in Luxembourgish texts
4. **Investigate symbol frequencies** and their optimal placement
5. **Propose placement strategies** for accented characters and symbols

For the full optimization process and comparison with other layouts (AZERTY, BÉPO, Dvorak),
see the companion vignette: `vignette("luxembourg-optimization")`.

---

## 1. Setup: The Multilingual Corpus

We use a balanced corpus representing Luxembourg's language landscape:

```{r load-corpus}
# Load sample texts
data("french")
data("german")
data("english")
data("luxembourguish")

# Create weighted corpus (30% FR, 30% EN, 20% DE, 20% LB)
corpus <- list(
  French = rep(french, 3),
  English = rep(english, 3),
  German = rep(german, 2),
  Luxembourgish = rep(luxembourguish, 2)
)

# Combine all texts
all_texts <- unlist(corpus)

# Calculate frequencies
freq_all <- letter_freq(paste(all_texts, collapse = " "))

# Display corpus stats
cat("Total corpus size:", format(sum(nchar(all_texts)), big.mark = ","), "characters\n")
cat("Unique characters:", nrow(freq_all), "\n")
```

### Top 15 Most Frequent Letters

```{r freq-table}
freq_top <- head(freq_all, 15)
freq_top$pct <- paste0(round(freq_top$frequencies * 100, 2), "%")

knitr::kable(
  freq_top[, c("characters", "total", "pct")],
  col.names = c("Letter", "Count", "Frequency"),
  caption = "Top 15 most frequent characters in the multilingual corpus"
)
```

---

## 2. The Two Layouts

### Standard QWERTZ (Swiss)

```{r qwertz-standard}
data("ch_qwertz")

# Extract only letter keys from the full keyboard layout
# ch_qwertz contains 119 keys; we need just the 26 letters
qwertz_letters <- ch_qwertz %>%
  filter(tolower(key) %in% letters) %>%
  arrange(row, number)

cat("Standard QWERTZ Layout:\n\n")
print_layout(qwertz_letters)
```

### Optimized QWERTZ with Home Row Optimization

The optimized layout keeps the familiar left-hand positions (A, S, D, E) while
placing the **6 next most frequent letters** on the remaining home row positions:

```{r qwertz-optimized, results='hide'}
# Fixed keys for home row: A, S, D, E (positions 0-3)
fixed_home <- c("a", "s", "d", "e")

# Get top frequent letters excluding the fixed ones
freq_letters <- as.character(freq_all$characters)
top_freq_excluding_fixed <- setdiff(freq_letters, fixed_home)

# Next 6 most frequent letters for home row positions 4-9
next_6 <- head(top_freq_excluding_fixed, 6)

# Home row: ASDE (fixed) + next 6 frequent letters
home_row_keys <- c(fixed_home, next_6)

# Remaining letters for top and bottom rows
remaining_letters <- setdiff(letters, home_row_keys)

# QWERTZ positions for reference
qwertz_top_row <- c("q", "w", "e", "r", "t", "z", "u", "i", "o", "p")
qwertz_bottom_row <- c("y", "x", "c", "v", "b", "n", "m")

# Keep remaining letters in their QWERTZ positions where possible
top_row_keys <- sapply(qwertz_top_row, function(k) {
  if (k %in% remaining_letters) k else NA
})
bottom_row_keys <- sapply(qwertz_bottom_row, function(k) {
  if (k %in% remaining_letters) k else NA
})

# Letters that need new positions (displaced from home row)
letters_needing_placement <- remaining_letters[!remaining_letters %in% c(top_row_keys, bottom_row_keys)]

# Fill NA positions with displaced letters
fill_idx <- 1
for (i in seq_along(top_row_keys)) {
  if (is.na(top_row_keys[i]) && fill_idx <= length(letters_needing_placement)) {
    top_row_keys[i] <- letters_needing_placement[fill_idx]
    fill_idx <- fill_idx + 1
  }
}
for (i in seq_along(bottom_row_keys)) {
  if (is.na(bottom_row_keys[i]) && fill_idx <= length(letters_needing_placement)) {
    bottom_row_keys[i] <- letters_needing_placement[fill_idx]
    fill_idx <- fill_idx + 1
  }
}

# Remove NAs
top_row_keys <- top_row_keys[!is.na(top_row_keys)]
bottom_row_keys <- bottom_row_keys[!is.na(bottom_row_keys)]

# Build the full keyboard layout
ext_keys_qwertz <- c(top_row_keys, home_row_keys, bottom_row_keys)

# Fixed keys: ASDE (first 4 home row positions) + letters that stayed in QWERTZ positions
keys_in_original_position <- c(
  intersect(top_row_keys, qwertz_top_row),
  intersect(bottom_row_keys, qwertz_bottom_row)
)
fixed_keys_qwertz <- c(fixed_home, keys_in_original_position)
movable_keys <- setdiff(ext_keys_qwertz, fixed_keys_qwertz)

# Create keyboard data frame
kb_qwertz_opt <- data.frame(
  key = ext_keys_qwertz,
  key_label = toupper(ext_keys_qwertz),
  row = c(rep(1, length(top_row_keys)), rep(2, 10), rep(3, length(bottom_row_keys))),
  number = c(seq_along(top_row_keys) - 1, 0:9, seq_along(bottom_row_keys) - 1),
  stringsAsFactors = FALSE
)
kb_qwertz_opt$x_mid <- kb_qwertz_opt$number + c(0, 0.25, 0.5)[kb_qwertz_opt$row]
kb_qwertz_opt$y_mid <- kb_qwertz_opt$row

# Effort weights
effort_weights <- list(
  base = 3.0,
  same_finger = 3.0,
  same_hand = 0.5,
  row_change = 0.5,
  trigram = 0.3
)

# Optimize the remaining movable keys
GENERATIONS <- 200
POPULATION_SIZE <- 80

result_qwertz_home <- optimize_layout(
  text_samples = all_texts,
  keyboard = kb_qwertz_opt,
  keys_to_optimize = ext_keys_qwertz,
  rules = list(fix_keys(fixed_keys_qwertz)),
  generations = GENERATIONS,
  population_size = POPULATION_SIZE,
  effort_weights = effort_weights,
  verbose = FALSE
)
```

```{r show-optimized}
cat("Optimized QWERTZ Layout (Home Row: ASDE +", paste(next_6, collapse = " "), "):\n\n")
print_layout(result_qwertz_home$layout)
```

### Side-by-Side Comparison

```{r side-by-side}
cat("┌─────────────────────────────────────────────────────────────────────┐\n")
cat("│                    LAYOUT COMPARISON                               │\n")
cat("├─────────────────────────────────────────────────────────────────────┤\n")
cat("│                                                                     │\n")
cat("│  Standard QWERTZ              Optimized QWERTZ                     │\n")
cat("│  ─────────────────            ──────────────────                   │\n")
cat("│                                                                     │\n")

# Get keys from both layouts (use qwertz_letters already extracted earlier)
std_keys <- toupper(qwertz_letters$key)
opt_keys <- toupper(result_qwertz_home$layout$key)

# Show rows side by side
cat("│  ", paste(std_keys[1:10], collapse = " "), "    ", 
    paste(opt_keys[1:10], collapse = " "), "   │\n")
cat("│   ", paste(std_keys[11:19], collapse = " "), "       ", 
    paste(opt_keys[11:19], collapse = " "), "     │\n")
cat("│    ", paste(std_keys[20:26], collapse = " "), "           ", 
    paste(opt_keys[20:26], collapse = " "), "       │\n")
cat("│                                                                     │\n")
cat("└─────────────────────────────────────────────────────────────────────┘\n")
```

**Key Differences:**

- Home row now contains the 10 most frequent letters
- Letters A, S, D, E remain in their familiar positions (left hand)
- High-frequency letters like N, T, R, I are moved to home row right hand

---

## 3. Typing Effort Comparison

```{r effort-comparison}
# Calculate effort for standard QWERTZ
effort_qwertz <- calculate_layout_effort(
  ch_qwertz, all_texts,
  keys_to_evaluate = letters,
  effort_weights = effort_weights
)

# Get effort for optimized layout
effort_optimized <- result_qwertz_home$effort

# Create comparison table
effort_df <- data.frame(
  Layout = c("Standard QWERTZ", "Optimized QWERTZ"),
  Effort = c(effort_qwertz, effort_optimized),
  stringsAsFactors = FALSE
)

effort_df <- effort_df %>%
  mutate(
    Relative = round(Effort / min(Effort) * 100, 1),
    Improvement = paste0(round((1 - Effort / max(Effort)) * 100, 1), "%")
  )

knitr::kable(
  effort_df,
  col.names = c("Layout", "Effort Score", "Relative (%)", "Improvement"),
  caption = "Typing effort comparison (lower is better)"
)
```

```{r effort-barplot, fig.height=3, fig.width=6}
ggplot(effort_df, aes(x = reorder(Layout, Effort), y = Effort, fill = Layout)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = round(Effort, 1)), vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("Standard QWERTZ" = "#e74c3c", 
                                "Optimized QWERTZ" = "#27ae60")) +
  labs(
    title = "Typing Effort Comparison",
    subtitle = "Lower effort = more ergonomic typing",
    x = NULL,
    y = "Effort Score"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(size = 11)
  ) +
  coord_cartesian(ylim = c(0, max(effort_df$Effort) * 1.15))
```

---

## 4. Heatmap Visualization

Heatmaps show where typing effort concentrates. Brighter/warmer colors indicate
more frequently used keys.

### Standard QWERTZ Heatmap

```{r heatmap-qwertz, fig.width=8, fig.height=4, warning=FALSE}
heatmap_qwertz <- heatmapize(ch_qwertz, freq_all)
ggkeyboard(heatmap_qwertz) + 
  ggplot2::ggtitle("Standard QWERTZ - Frequency Heatmap") +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5, face = "bold"))
```

**Observation:** High-frequency letters (E, N, T, R) are scattered across multiple rows.

### Optimized QWERTZ Heatmap

```{r heatmap-optimized, fig.width=8, fig.height=4, warning=FALSE}
# Create heatmap for optimized layout using QWERTZ as template
qwertz_home_kb <- ch_qwertz
letter_mask <- tolower(qwertz_home_kb$key) %in% letters
letter_indices <- which(letter_mask)

# Get the optimized layout keys
opt_keys_lower <- result_qwertz_home$layout$key
if (length(letter_indices) == length(opt_keys_lower)) {
  qwertz_home_kb$key[letter_indices] <- opt_keys_lower
  qwertz_home_kb$key_label[letter_indices] <- toupper(opt_keys_lower)
}

heatmap_optimized <- heatmapize(qwertz_home_kb, freq_all)
ggkeyboard(heatmap_optimized) + 
  ggplot2::ggtitle("Optimized QWERTZ - Frequency Heatmap") +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5, face = "bold"))
```

**Observation:** The home row is now "hot" — most frequent letters are concentrated 
where fingers naturally rest, reducing hand movement and effort.

---

## 5. Accented Character Frequency Analysis

Luxembourg's multilingual environment requires efficient access to accented characters.
Let's analyze their frequency in our corpus:

```{r accent-frequency}
# Define accented characters of interest
accented_chars <- c(
  # French accents
  "é", "è", "ê", "ë", "à", "â", "î", "ô", "û", "ç",
  # German accents
  "ä", "ö", "ü", "ß"
)

# Count occurrences in corpus
combined_text <- paste(all_texts, collapse = " ")
combined_lower <- tolower(combined_text)

accent_counts <- sapply(accented_chars, function(char) {
  # Count both uppercase and lowercase
  sum(
    lengths(gregexpr(char, combined_lower, fixed = TRUE)),
    lengths(gregexpr(toupper(char), combined_text, fixed = TRUE))
  )
})

# Adjust for -1 returns (no match)
accent_counts <- pmax(accent_counts, 0)

# Create dataframe
accent_df <- data.frame(
  Character = accented_chars,
  Count = accent_counts,
  stringsAsFactors = FALSE
) %>%
  arrange(desc(Count)) %>%
  mutate(
    Frequency = paste0(round(Count / nchar(combined_text) * 100, 3), "%"),
    Language = case_when(
      Character %in% c("ä", "ö", "ü", "ß") ~ "German",
      TRUE ~ "French"
    )
  )

knitr::kable(
  accent_df,
  col.names = c("Character", "Count", "Frequency", "Primary Language"),
  caption = "Accented character frequencies in the multilingual corpus"
)
```

```{r accent-barplot, fig.height=4, fig.width=8}
# Filter to only show accents with non-zero counts
accent_plot_df <- accent_df %>% filter(Count > 0)

if (nrow(accent_plot_df) > 0) {
  ggplot(accent_plot_df, aes(x = reorder(Character, Count), y = Count, fill = Language)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("French" = "#3498db", "German" = "#e74c3c")) +
    labs(
      title = "Accented Character Frequency",
      subtitle = "Characters appearing in the multilingual corpus",
      x = "Character",
      y = "Occurrences"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "bottom"
    )
}
```

### Priority Accented Characters

Based on frequency analysis, the **priority accented characters** for direct access are:

```{r priority-accents}
priority_accents <- accent_df %>%
  filter(Count > 0) %>%
  head(5)

if (nrow(priority_accents) > 0) {
  cat("High-priority accented characters for direct keyboard access:\n\n")
  for (i in seq_len(nrow(priority_accents))) {
    cat("  ", i, ". ", priority_accents$Character[i], 
        " (", priority_accents$Count[i], " occurrences)\n", sep = "")
  }
}
```

---

## 6. Symbol Frequency Analysis

Let's analyze the frequency of common symbols in our corpus:

```{r symbol-frequency}
# Define symbols of interest
common_symbols <- c(
  # Punctuation
  ".", ",", ";", ":", "!", "?", "'", "\"", "-", "(", ")", 
  # Additional
  "/", "@", "#", "&", "*", "+", "=", "<", ">", "[", "]", "{", "}"
)

# Count occurrences
symbol_counts <- sapply(common_symbols, function(sym) {
  count <- length(gregexpr(sym, combined_text, fixed = TRUE)[[1]])
  if (count == 1 && gregexpr(sym, combined_text, fixed = TRUE)[[1]][1] == -1) {
    return(0)
  }
  return(count)
})

# Create dataframe
symbol_df <- data.frame(
  Symbol = common_symbols,
  Count = symbol_counts,
  stringsAsFactors = FALSE
) %>%
  arrange(desc(Count)) %>%
  mutate(
    Frequency = paste0(round(Count / nchar(combined_text) * 100, 3), "%"),
    Category = case_when(
      Symbol %in% c(".", ",", ";", ":", "!", "?") ~ "Punctuation",
      Symbol %in% c("'", "\"", "-", "(", ")") ~ "Common",
      TRUE ~ "Technical"
    )
  )

# Show top symbols
knitr::kable(
  head(symbol_df, 15),
  col.names = c("Symbol", "Count", "Frequency", "Category"),
  caption = "Top 15 symbol frequencies in the multilingual corpus"
)
```

```{r symbol-barplot, fig.height=4, fig.width=8}
symbol_plot_df <- symbol_df %>% 
  filter(Count > 0) %>%
  head(12)

if (nrow(symbol_plot_df) > 0) {
  ggplot(symbol_plot_df, aes(x = reorder(Symbol, Count), y = Count, fill = Category)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c(
      "Punctuation" = "#9b59b6",
      "Common" = "#3498db",
      "Technical" = "#95a5a6"
    )) +
    labs(
      title = "Symbol Frequency",
      subtitle = "Most common symbols in the multilingual corpus",
      x = "Symbol",
      y = "Occurrences"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "bottom"
    )
}
```

---

## 7. Placement Strategies for Accented Characters and Symbols

Based on our analysis, we now investigate optimal placement strategies for frequently 
used accented characters and symbols.

### Available Positions

On a standard keyboard, accented characters can be placed:

1. **AltGr + Base Letter**: Most intuitive (e.g., AltGr+e = é)
2. **Dedicated Keys**: Replace rarely used keys on the main layer
3. **Number Row Shift Layer**: Replace symbols like @ or #
4. **Extended Layout**: Add additional keys (as done in `luxembourg-optimization.Rmd`)

### Recommended Placement

```{r placement-recommendations}
# Create recommendation table
placement_rec <- data.frame(
  Character = c("é", "è", "à", "ü", "ö", "ç", "ä", "ê"),
  Method = c(
    "AltGr + E",
    "AltGr + ` then E (dead key)",
    "AltGr + A",
    "AltGr + U",
    "AltGr + O",
    "AltGr + C",
    "AltGr + Shift + A",
    "AltGr + ^ then E (dead key)"
  ),
  Alternative = c(
    "Dedicated key on extended layout",
    "Direct key if very frequent",
    "Dedicated key on extended layout",
    "Dedicated key (German layout)",
    "Dedicated key (German layout)",
    "Dedicated key on extended layout",
    "Dedicated key (German layout)",
    "Keep as dead key combination"
  ),
  Priority = c("High", "Medium", "High", "Medium", "Medium", "High", "Low", "Low"),
  stringsAsFactors = FALSE
)

knitr::kable(
  placement_rec,
  col.names = c("Character", "Primary Method", "Alternative", "Priority"),
  caption = "Recommended accented character placement strategies"
)
```

### Symbol Placement Principles

For frequently used symbols:

| Symbol | Current Position | Recommendation |
|--------|-----------------|----------------|
| `.` (period) | Easy access | Keep as-is |
| `,` (comma) | Easy access | Keep as-is |
| `'` (apostrophe) | Shift layer | Consider promoting to main layer |
| `-` (hyphen) | Main layer | Keep as-is |
| `;` `:` | Shift layer | Acceptable position |

---

## 8. Summary and Recommendations

### Typing Effort Improvement

```{r summary-improvement}
improvement_pct <- round((1 - effort_optimized / effort_qwertz) * 100, 1)
cat("The optimized QWERTZ layout reduces typing effort by", improvement_pct, 
    "% compared to standard QWERTZ.\n")
```

### Key Findings

```{r key-findings, echo=FALSE, results='asis'}
findings <- data.frame(
  Aspect = c(
    "Home Row Efficiency",
    "Learning Curve",
    "Accented Characters",
    "Symbol Access"
  ),
  `Standard QWERTZ` = c(
    "Suboptimal - frequent letters on top/bottom rows",
    "None - familiar layout",
    "Dead keys or AltGr required",
    "Complex combinations for some symbols"
  ),
  `Optimized QWERTZ` = c(
    "Excellent - 10 most frequent letters on home row",
    paste0("Moderate - ", length(movable_keys), " keys to relearn"),
    "Can add dedicated keys for é, ç, à, ü, ö",
    "Same as QWERTZ (unchanged)"
  ),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

knitr::kable(findings, caption = "Comparison summary")
```

### Recommendations

| Use Case | Recommendation |
|----------|---------------|
| **General multilingual typing** | Optimized QWERTZ with home row optimization |
| **Need direct accent access** | Extend layout with dedicated accent keys (é, ç, à, ü, ö) |
| **Minimal relearning** | Standard QWERTZ with AltGr accent access |
| **Programming focus** | Consider symbol layer optimization in future work |

---

## 9. QWERTZ-LUX: A BÉPO-Inspired Layout for Luxembourg

Based on our frequency analysis and inspired by the BÉPO layout's multi-layer approach,
we now implement and compare **QWERTZ-LUX** (QWERTZ, BÉPO-inspired for Luxembourg).

### Design Principles

1. **Direct access to frequent accents**: é, ä, ë, ç on letter rows
2. **Optimized home row**: High-frequency letters in best positions
3. **Preserve keyboard shortcuts**: Y, X, C, V, B on bottom row for Ctrl+X/C/V
4. **Intuitive AltGr layer**: Less frequent accents via AltGr + base letter

### The QWERTZ-LUX Layout

```{r qwertz-lux-layout}
# Create QWERTZ-LUX keyboard with direct accent access
qwertz_lux <- create_qwertz_lux_keyboard()

cat("QWERTZ-LUX Layout (33 keys including é, ä, ë, ç):\n\n")
cat("┌─────────────────────────────────────────────────────────────────────────┐\n")
cat("│  TOP ROW:     Q   W   F   O   G   Z   U   K   L   P   J                │\n")
cat("├─────────────────────────────────────────────────────────────────────────┤\n")
cat("│  HOME ROW:    A   S   D   E   ,   H   N   T   R   I   M   É            │\n")
cat("│               ↑   ↑   ↑   ↑   ↑                       ↑                │\n")
cat("│              (familiar)  (comma)                 (direct!)             │\n")
cat("├─────────────────────────────────────────────────────────────────────────┤\n")
cat("│  BOTTOM ROW:  Y   X   C   V   B   .   Ä   '   Ë   Ç                    │\n")
cat("│               ↑   ↑   ↑   ↑   ↑       ↑       ↑   ↑                    │\n")
cat("│              (shortcuts)        (direct!)  (Lux!) (ç)                  │\n")
cat("└─────────────────────────────────────────────────────────────────────────┘\n")
```

### Accent Access Comparison

| Accent | Frequency | Standard QWERTZ | QWERTZ-LUX |
|--------|-----------|-----------------|------------|
| é | 1003 | AltGr (+20%) | **Direct** (0%) |
| ä | 500 | AltGr (+20%) | **Direct** (0%) |
| ë | 273 | AltGr (+20%) | **Direct** (0%) |
| ç | 5 | AltGr (+20%) | **Direct** (0%) |
| ü | 170 | AltGr (+20%) | AltGr (+20%) |
| è | 112 | AltGr (+20%) | AltGr (+20%) |
| ö | 36 | AltGr (+20%) | AltGr (+20%) |
| à | 77 | AltGr (+20%) | AltGr (+20%) |
| ê, ô, î, û, â | rare | Dead key (+40%) | Dead key (+40%) |

### Fair Effort Comparison

Now we compare all three layouts with proper accent penalties:

```{r fair-comparison}
# Keys to evaluate for each layout
letters_only <- letters
lux_keys <- c(letters, "é", "ä", "ë", "ç")

# 1. Standard QWERTZ (letters + accent penalties for ALL accents)
effort_qwertz_fair <- calculate_layout_effort(
  ch_qwertz, all_texts, 
  keys_to_evaluate = letters_only,
  layer_map = qwertz_accent_layers,
  effort_weights = effort_weights
)

# 2. Optimized QWERTZ (letters + accent penalties for ALL accents)
effort_optimized_fair <- calculate_layout_effort(
  result_qwertz_home$layout, all_texts, 
  keys_to_evaluate = letters_only,
  layer_map = qwertz_accent_layers,
  effort_weights = effort_weights
)

# 3. QWERTZ-LUX (optimized letters + direct accents + remaining accent penalties)
effort_lux <- calculate_layout_effort(
  qwertz_lux, all_texts, 
  keys_to_evaluate = lux_keys,
  layer_map = qwertz_lux_accent_layers,
  effort_weights = effort_weights
)

# Create comparison table
comparison_df <- data.frame(
  Layout = c("Standard QWERTZ", "Optimized QWERTZ", "QWERTZ-LUX"),
  Effort = c(effort_qwertz_fair, effort_optimized_fair, effort_lux),
  stringsAsFactors = FALSE
) %>%
  mutate(
    Relative = round(Effort / max(Effort) * 100, 1),
    Improvement = paste0(round((1 - Effort / max(Effort)) * 100, 1), "%")
  )

knitr::kable(
  comparison_df,
  col.names = c("Layout", "Total Effort", "Relative %", "Improvement"),
  caption = "Fair comparison including accent access penalties"
)
```

```{r comparison-barplot, fig.width=8, fig.height=4}
ggplot(comparison_df, aes(x = reorder(Layout, -Effort), y = Effort, fill = Layout)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = Improvement), vjust = -0.5, size = 4, fontface = "bold") +
  scale_fill_manual(values = c(
    "Standard QWERTZ" = "#E74C3C",
    "Optimized QWERTZ" = "#F39C12", 
    "QWERTZ-LUX" = "#27AE60"
  )) +
  labs(
    title = "Typing Effort Comparison",
    subtitle = "Including accent access penalties (AltGr: +20%, Dead keys: +40%)",
    x = NULL, y = "Total Effort"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

### Key Findings

```{r key-findings-lux, echo=FALSE}
improvement_opt <- round((1 - effort_optimized_fair / effort_qwertz_fair) * 100, 1)
improvement_lux <- round((1 - effort_lux / effort_qwertz_fair) * 100, 1)
lux_vs_opt <- round((1 - effort_lux / effort_optimized_fair) * 100, 1)
```

1. **QWERTZ-LUX is `r improvement_lux`% more efficient** than Standard QWERTZ
2. **Optimized QWERTZ is `r improvement_opt`% more efficient** than Standard QWERTZ
3. **QWERTZ-LUX is `r lux_vs_opt`% more efficient** than Optimized QWERTZ alone

The additional improvement of QWERTZ-LUX over Optimized QWERTZ comes from:

- **Direct access to é, ä, ë, ç** (no AltGr penalty for the most common accents)
- **Physical key positions** for accents instead of virtual layer access

### Recommendations

| Use Case | Recommended Layout |
|----------|-------------------|
| **Luxembourg multilingual typing** | **QWERTZ-LUX** (best overall) |
| **Minimal relearning** | Optimized QWERTZ (familiar layout) |
| **Programming focus** | Standard QWERTZ (familiar for shortcuts) |
| **Pure French typing** | Consider BÉPO (beyond this comparison) |

---

## Session Info

```{r session-info}
sessionInfo()
```
